---
title: "Untitled"
output: github_document
---

```{r}
library(tidycensus)
library(tidyverse)

```

```{r}
readRenviron("~/.Renviron")
Sys.getenv("1fe353bf7e60c7e3d19a1aa328b64ad3054a5201")

```

Selected Core Variables

- Total Population (B01003_001): For reference and normalization.
- Median Household Income (B19013_001): Indicator of economic conditions.
- Unemployment Rate (calculated): B23025_005 (Unemployed) / B23025_003 (Labor Force).
- Education Level (Bachelor's or Higher): Sum of B15003_022 to B15003_025.
- Racial Composition:
White Population Percentage: B02001_002 / B01003_001.
Black Population Percentage: B02001_003 / B01003_001.
- Poverty Rate: B17001_002 (Below Poverty) / B17001_001 (Total Population).
- Uninsured Rate: B27001_005 / B01003_001.
- Housing Vacancy Rate: B25002_003 (Vacant Units) / B25002_001 (Total Units).

```{r}
# Define the variables of interest
variables <- c(
  total_population = "B01003_001",
  median_income = "B19013_001",
  unemployed = "B23025_005",
  labor_force = "B23025_003",
  bachelors_degree = "B15003_022",
  masters_degree = "B15003_023",
  professional_school_degree = "B15003_024",
  doctorate_degree = "B15003_025",
  poverty = "B17001_002",
  vacant_units = "B25002_003",
  total_units = "B25002_001",
  uninsured_male_under_6 = "B27001_004",
  uninsured_male_6_to_17 = "B27001_007",
  uninsured_male_18_to_24 = "B27001_010",
  uninsured_male_25_to_34 = "B27001_013",
  uninsured_male_35_to_44 = "B27001_016",
  uninsured_male_45_to_54 = "B27001_019",
  uninsured_male_55_to_64 = "B27001_022",
  uninsured_male_65_to_74 = "B27001_025",
  uninsured_male_75_and_over = "B27001_028",
  uninsured_female_under_6 = "B27001_033",
  uninsured_female_6_to_17 = "B27001_036",
  uninsured_female_18_to_24 = "B27001_039",
  uninsured_female_25_to_34 = "B27001_042",
  uninsured_female_35_to_44 = "B27001_045",
  uninsured_female_45_to_54 = "B27001_048",
  uninsured_female_55_to_64 = "B27001_051",
  uninsured_female_65_to_74 = "B27001_054",
  uninsured_female_75_and_over = "B27001_057"
)

# Retrieve the data for states using the latest 5-year ACS
state_data <- get_acs(
  geography = "state",
  variables = variables,
  year = 2022,
  survey = "acs5",
  output = "wide"
)

# Calculate derived variables and reorder columns
state_data <- state_data %>%
  mutate(
    unemployment_rate = (unemployedE / labor_forceE) * 100,
    bachelors_and_higher = (bachelors_degreeE + masters_degreeE + professional_school_degreeE + doctorate_degreeE) / total_populationE * 100,
    poverty_rate = (povertyE / total_populationE) * 100,
    total_uninsured = rowSums(select(., starts_with("uninsured_"))),
    uninsured_rate = (total_uninsured / total_populationE) * 100,
    vacancy_rate = (vacant_unitsE / total_unitsE) * 100
  ) %>%
  select(NAME, total_populationE, median_incomeE, unemployment_rate, 
         bachelors_and_higher, 
         poverty_rate, uninsured_rate,vacancy_rate)

# View the dataset
head(state_data)

```








