---
title: 'Data Cleaning'
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

<br>

### Define the Data Cleaning Function
The `clean_data` function standardizes the dataset structure and formats the dates. It adapts column names based on the number of columns in the dataset and handles inconsistencies such as parentheses, brackets, and whitespace in the data.

- Rename columns based on the structure of the dataset.
- Handle differences in column counts (e.g., 6 or 7 columns).
- Remove unnecessary characters like parentheses and brackets.
- Trim whitespaces and clean special characters.
- Standardize and format dates, adding the year if missing.
```{r message=FALSE}
library(dplyr)
library(tidyr)
library(stringr)

# Function to clean and unify data
clean_data <- function(data, year) {
  # Rename columns based on the number of columns
  col_count <- ncol(data)
  
  if (col_count == 6) {
    # For 2018 data
    colnames(data) <- c("Date", "Community", "Dead", "Injured", "Total", "Description")
    data <- data %>%
      separate(Community, into = c("City", "State"), sep = ", ", fill = "right")
  } else if (col_count == 7) {
    # For other years
    colnames(data) <- c("Date", "City", "State", "Dead", "Injured", "Total", "Description")
  } else {
    stop("Unexpected number of columns in the data")
  }
  
  # Data cleaning
  data <- data %>%
    # Remove content in parentheses or brackets
    mutate(across(everything(), ~str_remove_all(., "\\(.*?\\)|\\[.*?\\]"))) %>%
    # Trim whitespace and remove special characters
    mutate(across(everything(), ~str_trim(.))) %>%
    mutate(Date = str_replace_all(Date, "[^a-zA-Z0-9,– ]", "")) %>%
    # Handle date ranges like "November 19–20" or "February 11–12"
    mutate(Date = if_else(str_detect(Date, "–"), 
                          str_extract(Date, "^[a-zA-Z]+ \\d+"), 
                          Date)) %>%
    # Add year to Date if missing and standardize format
    mutate(Date = ifelse(!str_detect(Date, "\\d{4}$"), paste0(Date, ", ", year), Date)) %>%
    # Convert to Date format
    mutate(Date = as.Date(Date, format = "%B %d, %Y"))
  
  return(data)
}
```
### Load and Clean Data for Each Year

The following steps are repeated for each year's dataset (2018-2024):

- Load the dataset for the given year.
- Clean the dataset using the clean_data function.
- Store the cleaned dataset in a list for further processing.
```{r}
# Load and clean each year's data
years <- 2018:2024
all_data <- list()

for (year in years) {
  
  file_path <- paste0("data/mass_shootings_", year, ".csv")
  data <- read.csv(file_path, stringsAsFactors = FALSE)
  
  # Clean and store
  cleaned_data <- clean_data(data, year)
  all_data[[as.character(year)]] <- cleaned_data
}
```
### Combine and Finalize the Data

After cleaning the data for all years:

- Combine the datasets into a single dataframe.
- Remove rows with missing values.
- Sort the dataset by date in descending order.
```{r}
# Combine all years into one dataframe
final_data <- bind_rows(all_data)

# Remove rows with any NA values
final_data <- final_data %>%
  filter(complete.cases(.))

# Sort the data in descending order by Date
final_data_sorted <- final_data %>%
  arrange(desc(Date))
```
### Save the Cleaned Data

Export the cleaned and combined dataset to a new CSV file for future use.
```{r}
write.csv(final_data_sorted, "data/mass_shootings_2018_2024_cleaned.csv", row.names = FALSE)
```





